A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-20 paper/writeup.tex
some edits

\newcommand*{\tabref}[2][]{%
	\hyperref[{#2}]{%
		Table~\ref*{#2}%
		\ifx\\#1\\%
		\else
		\,#1%
		\fi
	}%
}

@@

\author[$\dagger$]{Halley Fritze}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\ast$,$\dagger$,S]{Peter Ralph}

@@

Haplotype identity has been nearly overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure sharing of haplotypes between adjacent trees.

@@

Consider the (small portion) of a hypothetical tree sequence in \figref{fig:conceptual}A.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, things change once we think about what this implies about haplotype inheritance.
\figref{fig:conceptual}B shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in \figref{fig:conceptual}C,
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in \figref{fig:conceptual}D --
and so node 3 has become unary in this tree.
Note that the more parsimonious tree sequence also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref{fig:conceptual}B
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in \figref{fig:conceptual}D.

So, given the ARG shown in \figref{fig:conceptual}A\&B,
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref{fig:conceptual}C\&D,
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?

@@

We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
to take advantage of the extensive, stable, and well-documented set of open-source tools
available in \tskit{} \citep{tskit}.
So, our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
over a segment of genome $[\ell_e, r_e)$.
(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
for which the intermediate nodes are not in the other tree
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, there exists paths from root node $6$ to sample node $0$. $T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$. The intermediate nodes $4$ and $5$ do not appear in $T_2$ respectively $T_1$, and so we consider these two paths \textit{mergeable}. The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$. This new path extends the edge $4\to 0$ to $T_2$ as well as constructs a new edge $5\to 4$ in the tree sequence.
    \label{fig:extending_diagram}
}
\end{figure}

@@

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        checks to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
        $p_i = P_I[c]$\;
        $t_i = t[p_i]$\;
        $p_o = P_O[c]$\;
        $t_o = t[p_o]$\;
        $n_e = n = 0$\;
        \While{\tn{True}}{
            $y_i = (p_i \neq NULL)
                \text{ and } (p_i \notin T_O)
                \text{ and } (t_i < t_o)$ \;
            $y_o = (p_o \neq NULL)
                \text{ and } (p_o \notin T_I)
                \text{ and } (t_o < t_i)$ \;
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
                    $n_e += 1$\;
                }
                $c = p_i$\;
                $p_i = P_I[p_i]$\;
                $t_i = (p_i = NULL) ? \infty : t[p_i]$\;
            }{
                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
                    $n_e += 1$\;
                }
                $c = p_o$\;
                $p_o = P_O[p_o]$\;
                $t_o = (p_o = NULL) ? \infty : t[p_o]$\;
                $n += 1$\;
            }
        }
        \If{$n = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            $n_e = \infty$\;
        }
        \KwRet{$n_e$}\;
    }
\end{algorithm2e}

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes.
        \comment{TODO: include this in the previous algorithm?}
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
        \For{$k$ in $1 \ldots N$}{
            $M = 0$\;
            $M' = \infty$\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$)\;
                    \eIf{$m < M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
                    }{
                        $M' = \min(m, M)$\;
                    }
                }
                $M = M'$\;
                $M' = \infty$\;
            }
        }
    }
\end{algorithm2e}

@@

It is helpful to first describe what we compute in the simple case.
\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
for each node, the top point is the total span in a ``true'' tree sequence
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
The bottom point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
\comment{Usually (?)} between the two points
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
So, the difference between the middle and bottom points is the amount by which
the span of that node has been extended.
\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
for each node, it gives the percent of the span after running \algorithmref{alg:extend}
that is incorrect (i.e., not present in the original tree sequence).
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are not given an identification between nodes otherwise
(as for instance with a true, simulated tree sequence and one inferred from its genotypes).
\comment{should have introduced ``samples'' before this}
Call the two tree sequence $T_1$ and $T_2$, which should have the same genome length
and the same set of sample nodes.
We would like to measure (a) how much of $T_1$ is found in $T_2$;
(b) how much of $T_1$ is \emph{not} found in $T_2$; and
(c) how much of $T_2$ is not found in $T_1$.
(Think of these three quantities as the size of the intersection
and two differences between the tree sequences,
thought of vaguely as sets.)
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).

@@

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align*}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(\log t^{(1)}_n - \log t^{(2)}_{\alpha(n)} \right)^2 
    }{
        \sum_{n \in N_1} s(\T_1,n)
    } } .
\end{align*}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.

We compute the dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.

@@

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity function}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the dissimilarity $\dis(T1,T2)=\frac{3}{46}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

It's also interesting to  note that inference
(see \figref{fig:inferred_edge_counts})
(a) increases number of edges;
(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
(c) but then extending reduces to below the original.

@@

>>>>

\newcommand*{\tabref}[2][]{%
	\hyperref[{#2}]{%
		Table~\ref*{#2}%
		\ifx\\#1\\%
		\else
		\,#1%
		\fi
	}%
}
\newcommand*{\eqnref}[2][]{%
	\hyperref[{#2}]{%
		Equation~\ref*{#2}%
		\ifx\\#1\\%
		\else
		\,#1%
		\fi
	}%
}

@@

\author[$\dagger$]{Halley Fritze}
\author[$\ast$]{Nathaniel Pope}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\ast$,$\dagger$,S]{Peter Ralph}

@@

Haplotype identity has been nearly overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure sharing of haplotypes between adjacent trees.

@@

Consider the (small portion) of a hypothetical tree sequence in \figref[A]{fig:conceptual}.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, things change once we think about what this implies about haplotype inheritance.
\figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in \figref[C]{fig:conceptual},
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in \figref[D]{fig:conceptual} --
and so node 3 has become unary in this tree.
Note that the more parsimonious tree sequence also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref[B]{fig:conceptual}
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in \figref[D]{fig:conceptual}.

So, given the ARG shown in \figref[A\&B]{fig:conceptual},
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?

@@

We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
to take advantage of the extensive, stable, and well-documented set of open-source tools
available in \tskit{} \citep{tskit}.
So, our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Some nodes are \emph{samples}, meaning that we know them from data. 
Typically, samples are exclusively leaves of $T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
over a segment of genome $[\ell_e, r_e)$.
(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$,
for which the intermediate nodes are not in the other tree.
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, there exists paths from root node $6$ to sample node $0$. $T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$. The intermediate nodes $4$ and $5$ do not appear in $T_2$ and $T_1$ respectively, and so we consider these two paths \textit{mergeable}. The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$. This new path extends the edge $4\to 0$ to $T_2$ as well as constructs a new edge $5\to 4$ in the tree sequence.
    \label{fig:extending_diagram}
}
\end{figure}

@@

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
        $p_i = P_I[c]$\;
        $t_i = t[p_i]$\;
        $p_o = P_O[c]$\;
        $t_o = t[p_o]$\;
        $n_e = n = 0$\;
        \While{\tn{True}}{
            $y_i = (p_i \neq NULL)
                \text{ and } (p_i \notin T_O)
                \text{ and } (t_i < t_o)$ \;
            $y_o = (p_o \neq NULL)
                \text{ and } (p_o \notin T_I)
                \text{ and } (t_o < t_i)$ \;
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
                    $n_e += 1$\;
                }
                $c = p_i$\;
                $p_i = P_I[p_i]$\;
                $t_i = (p_i = NULL) ? \infty : t[p_i]$\;
            }{
                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
                    $n_e += 1$\;
                }
                $c = p_o$\;
                $p_o = P_O[p_o]$\;
                $t_o = (p_o = NULL) ? \infty : t[p_o]$\;
                $n += 1$\;
            }
        }
        \If{$n = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            $n_e = \infty$\;
        }
        \KwRet{$n_e$}\;
    }
\end{algorithm2e}

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{Extend haplotypes.}
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
        \For{$k$ in $1 \ldots N$}{
            $M = 0$\;
            $M' = \infty$\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$)\;
                    \eIf{$m < M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
                    }{
                        $M' = \min(m, M)$\;
                    }
                }
                $M = M'$\;
                $M' = \infty$\;
            }
        }
    }
\end{algorithm2e}

@@

It is helpful to first describe what we compute in the simple case.
\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
for each node, the dark blue point is the total span in a ``true'' tree sequence
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
The green point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
The red point
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
\figref[B]{fig:node-spans} shows how much of this span is incorrect:
for each node, it gives the amount of span
that is incorrect (i.e., not present in the original tree sequence)
over the total amount of span added after running \algorithmref{alg:extend}.
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are not given an identification between nodes otherwise
(as for instance with a true, simulated tree sequence and one inferred from its genotypes).
Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
and the same set of sample nodes.
We would like to measure (a) how much of $T_1$ is found in $T_2$;
(b) how much of $T_1$ is \emph{not} found in $T_2$; and
(c) how much of $T_2$ is not found in $T_1$.
(Think of these three quantities as the size of the intersection
and two differences between the tree sequences,
thought of vaguely as sets.)
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).

@@

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align}\label{eqn:wrmse}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(\log\left(1+ t^{(1)}_n\right) - \log\left(1+ t^{(2)}_{\alpha(n)}\right) \right)^2 
    }{
        \sum_{n \in N_1} s(\T_1,n)
    } } .
\end{align}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.

The implementation of this method, called \emph{tree\_dissimilarity} \comment{change?}, in \texttt{tsdate}
produces three values after computation: 
the relative dissimilarity, weighted root-mean-square-error, and true proportion represented. 
The \emph{relative dissimilarity} is defined to be the dissimilarity proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\overline{\dis}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}
 	=1-\frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}.
\end{align}
The \emph{weighted root-mean-square-error} is given in \eqnref{eqn:wrmse}.
The \emph{true proportion represented} is similarity between two trees relative 
to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\overline{\similarity}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_2}s(\T_2,n)}.
\end{align}
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
We compute the tree\_dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.

@@

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the dissimilarity $\dis(T1,T2)=3$ and the \emph{relative similarity} $\overline{\dis}(T1,T2)=\frac{3}{46}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

It's also interesting to  note that inference
(see \figref{fig:inferred_edge_counts})
(a) increases number of edges;
(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
(c) but then extending reduces to below the original.
\comment{Also say something about IE line...}

@@

\begin{figure}
	\includegraphics{newplots_wo_ee/figure-4b-supplement}
	\caption{Distribution of incorrectly added span percentages
		 for the simulated tree sequence in \figref{fig:node-spans}.}
	\label{fig:incorrect_ratio}
\end{figure}
